Gradual Soundness: Lessons from Static Python

### 2021-10-05 ecoop-2022/paper.tex
ecoop: intro

\documentclass[a4paper,english,cleveref,autoref,thm-restate,,]{lipics-v2021}
\bibliographystyle{plainurl}

@@

@@

%% TODO before camera-ready
%% -

@@

% > Here’s the thing: I think this is the sort of thing an academic might say “oh,
% > this is too much work, it’d never work” But in fact it looks like they seem to
% > be making it work
% -- Shriram 2021-09-30

\author{Kuang-Chen Lu}{Department of Computer Science, Brown University, USA}{LuKuangchen1024@gmail.com}{}{}
\author{Ben Greenman}{Department of Computer Science, Brown University, USA}{benjamin.l.greenman@gmail.com}{0000-0001-7078-9287}{}
\author{Carl Meyer}{Facebook, Inc.}{carljm@fb.com}{}{}
\author{Dino Viehland}{Facebook, Inc.}{dinoviehland@fb.com}{}{}
\author{Shriram Krishnamurthi}{Department of Computer Science, Brown University, USA}{joanrpublic@dummycollege.org}{0000-0001-5184-1975}{}

@@

@@

Go go gradual~\cite{tf-dls-2006,st-sfp-2006}.

@@

\begin{itemize}
\item
  Evidence that gradual soundness works. Performance rises incrementally from runtime engineering and programmer buy-in.
\item
  Informal description of the programming methods that enabled good performance.
  % migration strategy, use of dyn., use of handwritten annotations
\item Formal description of the core language, to promote a reproduction.
\end{itemize}

The language design is a novel point in the space, combining concrete and transient checking strategies.
Concrete has the best performance, but imposes an annotation burden.
Transient mitigates the burden.

@@

The dynamic type is an important part of Static Python.
Dynamic avoids the need for whole-sale migration,
lets programmers work around dynamic features like splatting,
and is (proof?) in some cases necessary for optimal performance.

>>>>

\documentclass[a4paper,english,cleveref,autoref,thm-restate,anonymous,]{lipics-v2021}
\bibliographystyle{plainurl}

@@

% > Here’s the thing: I think this is the sort of thing an academic might say “oh,
% > this is too much work, it’d never work” But in fact it looks like they seem to
% > be making it work -- Shriram 2021-09-30

@@

%% TODO after submission, before camera-ready
%% -

@@

\author{Kuang-Chen Lu}{Department of Computer Science, Brown University, USA}{LuKuangchen1024@gmail.com}{}{}
\author{Ben Greenman}{Department of Computer Science, Brown University, USA}{benjamin.l.greenman@gmail.com}{0000-0001-7078-9287}{}
\author{Carl Meyer}{Facebook, Inc.}{carljm@fb.com}{}{}
\author{Dino Viehland}{Facebook, Inc.}{dinoviehland@fb.com}{}{}
\author{Shriram Krishnamurthi}{Department of Computer Science, Brown University, USA}{shriram@brown.edu}{0000-0001-5184-1975}{}

@@

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}

@@

Gradual typing has attracted significant interest as a solution to
the old tradeoff between static and dynamic typing.
The premise is simple: let programmers apply types to part of the
codebase and leave the rest un-analyzed.
Run-time checks can in principle enforce the assumptions that
typed code makes about untyped components, thereby ensuring sound types.

Unfortunately, the high run-time cost of sound types has split
the community.
Industry teams have developed innovative type systems that accommodate
untyped designs, but these systems are all unsound [CITE].
They intentionally check nothing at run-time when untyped values enter
typed code.
Academic teams have primarily focused on the theory of sound
graduality and its extension to new types and correctness theorems [CITE].
A few academics have studied the cost of run-time checks
in detail [CITE] and proposed implementation methods [CITE],
compiler technology [CITE],
and even weakened semantics [CITE], but none of these efforts
decisively close the performance gap.
The most promising attempt is the \emph{concrete} semantics
for gradual types [CITE kafka nom etc], but its low performance
overhead stems from severe restrictions on untyped code.
These restrictions may not be acceptable to the maintainers of a
working untyped codebase.
Indeed, the maintainers might prefer to re-implement their system
in a typed language as Dropbox recently did to its core sync
engine.\footnote{\shorturl{https://}{dropbox.tech/infrastructure/rewriting-the-heart-of-our-sync-engine}}

In short, academic researchers are working to close the performance gap
without overly restricting the promise of gradual typing.
Industry researchers are, for the most part, sidestepping the problem with unsound types.

This paper reports on a remarkable exception to the rule among industry-made gradual type systems.
The \SP{} team at Instagram has developed a sound type system for a subset of
Python along with a runtime system designed to lower the cost of run-time checks
and capitalize on the benefits of sound types.
The language design is a novel twist on concrete types that lets programmers
begin with a shallow claims about data structures and gradually opt in to deep soundness.
Remarkably, the performance of Python is an upper bound on their observed performance of
Static Python.

@@

\begin{itemize}
  \item
    Evidence that gradual soundness can work.
    \SP{} performs well on a microbenchmark suite and, more importantly, in production.
    The effort needed to convert code to \SP{} is low thanks to its novel application
    of concrete types.
  \item
    A formal description and analysis of the \SP{} core language.
    The analysis includes a proof of soundness and (perhaps) a metric for
    the relative cost of a type migration.
  \item
    A full list of the limitations in \SP{} relative to idealized gradual type
    systems and an informal description of the programming discipline that Instagram
    follows in their gradually sound codebase.
\end{itemize}

\paragraph*{Significance}

We have written this paper to promote two future directions.
First, we want to encourage system-builders to reproduce the essentials of
\SP{} in another context and measure the net performance.
Second, we want to lend focus to theoretical research on gradual typing.
On one hand, the restrictions that \SP{} adopts (relative to the ideal promise of gradual typing)
may be worthwhile assumptions for theory going forward.
On the flip side, there are issues in today's \SP{} that might
be addressed by future research.

@@

The dynamic type is an important part of Static Python.
Dynamic avoids the need for whole-sale migration,
lets programmers work around dynamic features like splatting,
and is in some cases necessary for optimal performance (evidence?).
